A Solution to Static vs. Dynamic Linking

### 2024-04-25 mss.tex
Final cut, maybe pending bold font.

\begin{abstract}
\noindent
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  The Urbit \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

The problem with static linking is that if the same library is used by more than one application, there is more than one copy of it in memory as a result.  This memory duplication can be demanding on RAM utilization and reduce cache locality, degrading overall system performance.

Dynamic linking was invented to address this problem.  Instead of linking a library into a program at build time, one links it at runtime.  The OS keeps a single copy of the shared library in RAM that multiple programs can use.  This reduces memory usage and improves performance, but can lead to version mismatches and dependency issues.  \citeauthor{Locklin2022}~explains,

@@

Linking, in the Urbit sense, derives from supplying nouns to nouns at compile time.\footnote{This is perhaps the biggest distinction from the conventional scenario for linking, which refers to RAM words.  \emph{Ceteris paribus}, this article's discussion can inform the traditional dialogue.}  In the Urbit build arm \lstinline[style=inlinecode]{++ford}, a pair of builds becomes the build of a pair.  The subject (environment) used to build a file is the tuple \lstinline[style=inlinecode]{[import_n … import_2 import_1 stdlib]}.  Since Hoon symbol lookup is left-to-right, this nests scopes seamlessly and predictably.  The linking technique is essentially trivial.

@@

But how does one know whether the cached library is valid?  To achieve global (cross-application) deduplication, one needs a referentially transparent build cache, i.e. a build factored as a full description of an input to the build system.  Since the build system is deterministic, if one sees the same input, one knows that it will produce the same output.

@@

\section{The Modern \lstinline[style=inlinecode]{++ford} Build Cache}

\sloppy
The former Ford cache was per-desk,\footnote{Clay desks in Urbit are like Git branches, describing particular filesystem continuities.} keyed on the name of the build (e.g. a file at a certain path).  It was impossible to share such a cache between desks because the name may refer to different things on different desks or at different revisions.

>>>>

\begin{abstract}
\noindent
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  Urbit's \lstinline[style=inlinecode]{++ford} build system elegantly solves the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

The problem with static linking is that if the same library is used by more than one application, there is more than one copy of it in memory as a result.  This memory duplication can be demanding on {\sc ram} utilization and reduce cache locality, degrading overall system performance.

Dynamic linking was invented to address this problem.  Instead of linking a library into a program at build time, one links it at runtime.  The OS keeps a single copy of the shared library in {\sc ram} that multiple programs can use.  This reduces memory usage and improves performance, but can lead to version mismatches and dependency issues.  \citeauthor{Locklin2022}~explains,

@@

Linking, in the Urbit sense, derives from supplying nouns to nouns at compile time.\footnote{This is perhaps the biggest distinction from the conventional scenario for linking, which refers to {\sc ram} words.  \emph{Ceteris paribus}, this article's discussion can inform the traditional dialogue.}  In the Urbit build arm \lstinline[style=inlinecode]{++ford}, a pair of builds becomes the build of a pair.  The subject (environment) used to build a file is the tuple \lstinline[style=inlinecode]{[import_n … import_2 import_1 stdlib]}.  Since Hoon symbol lookup is left-to-right, this nests scopes seamlessly and predictably.  The linking technique is essentially trivial.

@@

But how does one know whether the cached library is valid?  To achieve global (cross-application) deduplication, one needs a referentially transparent build cache, i.\,e.\ a build factored as a full description of an input to the build system.  Since the build system is deterministic, if one sees the same input, one knows that it will produce the same output.

@@

\section[The Modern \texttt{++ford} Build Cache]{The Modern \texttt{++ford} Build Cache}

\sloppy
The former Ford cache was per-desk,\footnote{Clay desks in Urbit are like Git repositories, describing particular filesystem continuities.} keyed on the name of the build (e.\,g.\ a file at a certain path).  It was impossible to share such a cache between desks because the name may refer to different things on different desks or at different revisions.
